Stack Safety as a Security Property

### 2020-07-10 main.tex
Sections 1-2 pass.

The call stack is a perennial target for attacks
based on memory safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \ifaftersubmission\apt{Mostly from
  nick; there could be more}\leo{Do we want to hunt for more refs here?}
\bcp{Sure, but later!}
\fi

@@

The intuition behind \citeauthor{Skorstengaard+19}'s {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should not be able to access or overwrite private data
in the caller's stack frame.
%
We formalize this intuition in (1) a {\em stack confidentiality}
property, which promises that a callee's observable behavior is not affected
by memory outside its stack frame (e.g., a caller's private data or
uninitialized memory), plus (2) a {\em stack integrity} property,
which promises that, if a function call ever returns, the caller's stack
data at the point of return will be identical to what it was before the
call.

@@

After formally defining stack safety, we show that the definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider
\citeauthor{DBLP:conf/sp/RoesslerD18}'s
conservative Depth Isolation policy, which not only enforces our basic
stack safety property, but also an alternative lockstep property that
checks for integrity and confidentiality violations at every step of
an execution trace. The advantage of this property is that it ``fails
faster'' and is therefore more suitable for efficient testing. We prove (in
Coq) that this ``lockstep'' property implies our original definition of stack
safety.

The downside of the Depth Isolation policy is that it is rather
expensive to implement. In the same work,
\citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement.
Instead of enforcing the integrity and confidentiality of the stack at the
exact moment
when a callee returns, this policy signals violations when the
caller actually accesses data that the callee has written into its stack
frame\ifaftersubmission\bcp{We could also hint at the lazy confidentiality
  property in the same way...}\fi.
The lazy policy does not enforce our original integrity and confidentiality
properties, and to our knowledge there is no formal statement of the properties that
it ought to enforce. We propose ``observable'' versions of integrity and
confidentiality property, which capture the desired behavior of a policy that
detects violations between when they occur and when they make a visible impact.
Unfortunately the existing Lazy Tagging and Clearing policy of
\citeauthor{DBLP:conf/sp/RoesslerD18} does not quite enforce these properties,
but would if enhanced (at some performance cost).
%\bcp{I worry that we are being a little
%  dishonest here: the stated advantage of the lazy policy is that it is more
%efficient, but our variant is {\em not} efficient!}

\sna{There's been some inconsistency in ``stack safety'' vs.
     ``stack-safety,'' I'm fixing on the former.}
\bcp{Be careful: When a compound phrase is used as an adjective to modify a
  following noun, it gets hyphenated; otherwise it does not.  This is
  correct grammar, not inconsistency!  :-)}

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for this---''quasi-extensional''?}
  A novel {\em stack safety property} abstractly capturing the
  intuitive concepts of strict local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}).
  % These properties rule out common examples of stack memory abuse.
  This property can be enforced by
  the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, which is
  based on hardware tagging~(\cref{sec:conservative}).
\item
  A {\em lockstep} variant of the stack safety property that is also
  enforced by the Depth Isolation policy
  but is more efficient to test, plus a proof (in Coq) that this variant is
  stronger than the original~(\cref{sec:lockstep}).
\item
  A {\em lazy} variant of the stack safety property that can be enforced if the
  Lazy Tagging and Clearing policy of \citet{DBLP:conf/sp/RoesslerD18} is enhanced
  with unique tags on each activation, plus a proof (in Coq) that this
  variant is weaker than the original~(\cref{sec:lazy}).
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

@@

To make our stack safety definitions precise, we make some straightforward
assumptions about
memory layout; our definitions could easily be changed to be parametric over
these assumptions, at the cost of some complexity in their statements.  Memory
consists of an instruction region and a stack that grows upwards. (For simplicity,
we do not consider the heap. Integrating a memory safety property for the
heap in the style of
\citet{DBLP:conf/post/AmorimHP18} is a natural extension that we discuss
in~\Cref{sec:relwork}.)  The stack has a conventional structure, with
contiguously allocated per-function frames.
The boundary between the frames of a caller and callee is indicated by the $\rsp$ register
at the time of the call.  The stack can be used to
pass arguments or return results in the usual way, by placing them
at the top of the caller's frame. Callers and callees can also communicate
via registers if they wish; indeed, our 
definitions say nothing at all about registers (except $\PCname$ and
$\rsp$). \bcp{This implicitly amounts to choosing a caller-saves register
  convention, though, doesn't it?\apt{I think it is neutral; we could add separate caller register protection if we wanted.}}

Our stack safety concepts depend critically on the notions of ``call'' and ``return,''
which are not fully explicit in machine code (the same instruction opcode may be
used in some places as part of a call or return sequence and in other places
for other purposes).  We define a call to be
any instruction declared as such in the compiler-provided call structure.
If a call occurs at $\PCname = a$ and $\rsp = s$, then the \emph{corresponding
  return} has occurred when the $\PCname$ points to the next instruction after $a$ and $\rsp = s$. (Our definition
of well-bracketed control flow relates this definition of return to the program's
declared set of return points.)
Our property definitions make no other detailed assumptions about call or return sequences, although
some of the enforcement mechanisms we consider do.

@@

\paragraph*{Software-only policies}
%
Some policies are implemented in software, by modifying code in the
compiler, for example to perform bounds checking~\citep{NagarakatteZMZ09} or
to insert stack canaries~\citep{Cowan+98}. Such approaches should
still satisfy our formal notion of stack safety \apt{Stack canaries certainly would not!} with a trivial null
policy, but only for programs produced by the compiler.\apt{This still seems wrong to me. The policy part needs somehow to describe everything that isn't varied or rolled back.}
%These approaches are represented as policies that
%accept\bcp{what does it mean for a policy to accept a state??} only initial machine states that are well-formed for the policy (due to
%being produced by the compiler.) The set of policy states is again the unit type
%and the step function constant. Property enforcement comes from the fact that
%only ill-formed states can induce a trace that violates the property.

@@

In this section we examine how an existing enforcement mechanism,
Depth Isolation from \citet{DBLP:conf/sp/RoesslerD18}, can implement
the formal stack safety property described in the previous section: it in
fact enforces a stronger, inductive version of that property which strengthens the
local state encapsulation component by checking for integrity and
confidentiality violations at every step of the callee's execution trace.

@@

Our micropolicy works as follows. (We assume for ease of exposition that no
arguments are passed on the stack.)
%% \rb{Come back to this later?}.
The micropolicy maintains tags on values in memory of the form $\tagStackDepth{n}$,
indicating locations that belong to the stack frame at activation depth $n$,
or $\tagNoDepth$; it tags the $\PCname$ with tags of the form $\tagPCDepth{n}$.
During normal execution, the micropolicy rules only permit load and store operations
when the target memory is tagged with the same depth as the current {\PCname} tag.
Initially, the entire stack is tagged $\tagNoDepth$,
%\rb{actually, the ``initial frame'' should be tagged with its proper depth}
and the {\PCname} has tag $\tagPCDepth{0}$. These tags are altered
at exactly those points in
the program where the contour of the stack safety property changes:

@@

There remains the question of how to ensure that the sequences are
invoked at the right places. For a program to enjoy the stack safety property
(with respect to a particular call map), all we need require is that
the entry sequence be initiated at any instruction marked as a call
in the call map. This is achieved by giving these instructions (another) special tag,
and it is easy to check statically that this has been done correctly.
If the code fails to initiate an exit
sequence at a point where the stack safety property expects a return,
the micropolicy will incorrectly behave as if execution
is continuing in the callee, but since the callee never has stronger access
rights than the caller, this is harmless.

@@

This conservative policy satisfies not only the stack safety policy of
\cref{sec:lse-and-wbcf},
but also a provably (in Coq) stronger one that enforces local state
encapsulation at every point during execution. To encode that
property, we not only keep track of the execution of one machine
(whose trace we are trying to characterize as safe), but of an entire
stack of machine variations corresponding to nested calls of the
original machine. For each step of the original machine, we take a
step in every variant machine (thus, keeping them in lockstep
execution) and enforce a single-step variant of integrity and
confidentiality tests. In addition, every time a call is made we push
a variation of the original machine state at the time of that call on
top of the variant stack (just like the stack confidentiality property
enforced trace confidentiality for every call transition), and every time
the machine returns from a call we pop the top variation (to account
for stack confidentiality taking the prefix of the transition trace
until its return).

@@

  Part of the motivation for the strong lockstep stack safety property
  is that it uncovers errors quickly, and therefore should be quicker
  to test, just like the smarter testing properties
  of~\citet{TestingNI:ICFP}. We have already developed a testing
  framework for debugging our policy enforcement for this property, and
  it would be interesting to explore whether a testing-amenable variant
  of the lazy property also exists.

>>>>

The call stack is a perennial target for attacks
based on memory-safety vulnerabilities, with potential consequences
ranging from leakage or corruption of private stack data to
control-flow hijacking. To detect or prevent such attacks, a menagerie of
software and hardware techniques for stack safety have been proposed,
%
including stack canaries~\citep{Cowan+98},
bounds checking~\citep{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\citep{Kuznetsov+14},
shadow stacks~\citep{Dang+15,Shanbhogue+19},
capabilities~\citep{Woodruff+14,Chisnall+15,Skorstengaard+19,Skorstengaard+19b},
and hardware tagging~\citep{DBLP:conf/sp/RoesslerD18}. \ifaftersubmission\apt{Mostly from
  nick; there could be more}\leo{Do we want to hunt for more refs here?}
\bcp{Sure, but later!}
\fi

@@

The intuition behind \citeauthor{Skorstengaard+19}'s {\em
  local state encapsulation} is that callers should be protected from
callees: a callee should not be able to access or overwrite private data
in the caller's stack frame.
%
We formalize this intuition in (1) a {\em stack-confidentiality}
property, which promises that a callee's observable behavior is not affected
by memory outside its stack frame (e.g., a caller's private data or
uninitialized memory), plus (2) a {\em stack-integrity} property,
which promises that, if a function call ever returns, the caller's stack
data at the point of return will be identical to what it was before the
call.

@@

After formally defining stack safety, we show that the definition is
enforceable using an existing tag-based enforcement
mechanism~\citep{DBLP:conf/sp/RoesslerD18}. We first consider
\citeauthor{DBLP:conf/sp/RoesslerD18}'s
conservative Depth Isolation policy, which not only enforces our basic
stack-safety property, but also an alternative lockstep property that
checks for integrity and confidentiality violations at every step of
an execution trace. The advantage of this property is that it ``fails
faster'' and is therefore more suitable for efficient testing. We prove (in
Coq) that this ``lockstep'' property implies our original definition of stack
safety.

The downside of the Depth Isolation policy is that it is rather
expensive to implement. In the same work,
\citeauthor{DBLP:conf/sp/RoesslerD18} propose a different mechanism
that is more efficient by virtue of being lazy in its enforcement.
Instead of enforcing the integrity and confidentiality of the stack at the
exact moment
when a callee returns, this policy signals violations when the
caller actually accesses data that the callee has written into its stack
frame\ifaftersubmission\bcp{We could also hint at the lazy confidentiality
  property in the same way...}\fi.
The lazy policy does not enforce our original integrity and confidentiality
properties, and to our knowledge there is no formal statement of the properties that
it ought to enforce. We propose ``observable'' versions of integrity and
confidentiality property, which capture the desired behavior of a policy that
detects violations between when they occur and when they make a visible impact.
Unfortunately \citeauthor{DBLP:conf/sp/RoesslerD18}'s existing ``Lazy Tagging
and Clearing'' policy does not quite enforce these properties,
but would if enhanced (at some performance cost).
%\bcp{I worry that we are being a little
%  dishonest here: the stated advantage of the lazy policy is that it is more
%efficient, but our variant is {\em not} efficient!}

\sna{There's been some inconsistency in ``stack safety'' vs.
     ``stack-safety,'' I'm fixing on the former.}
\bcp{Be careful: When a compound phrase is used as an adjective to modify a
  following noun, it gets hyphenated; otherwise it does not.  This is
  correct grammar, not inconsistency!  :-)}
\sna{Aha, I forgot that. Going hard the other way.}

In summary, we offer the following contributions:
\begin{itemize}
\item
  % \leo{More? Extensional}\bcp{We might need to invent a word for this---''quasi-extensional''?}
  A novel {\em stack-safety property} abstractly capturing the
  intuitive concepts of strict local state encapsulation and well-bracketed
  control flow~(\cref{sec:lse-and-wbcf}).
  % These properties rule out common examples of stack memory abuse.
  This property can be enforced by
  the Depth Isolation policy of \citet{DBLP:conf/sp/RoesslerD18}, which is
  based on hardware tagging~(\cref{sec:conservative}).
\item
  A {\em lockstep} variant of the stack-safety property that is also
  enforced by the Depth Isolation policy
  but is more efficient to test, plus a proof (in Coq) that this variant is
  stronger than the original~(\cref{sec:lockstep}).
\item
  A {\em lazy} variant of the stack-safety property that can be enforced if the
  Lazy Tagging and Clearing policy of \citet{DBLP:conf/sp/RoesslerD18} is enhanced
  with unique tags on each activation, plus a proof (in Coq) that this
  variant is weaker than the original~(\cref{sec:lazy}).
%  This policy does not enforce our strict
%  local state encapsulation property,
%  but a modified version of it using unique per-activation tags enforces
%  an (also provably) weaker lazy variant of the property, which still prohibits
%  the same intuitively stack unsafe behaviors.
%
% \apt{Awkward, but the truth I fear.} \bcp{I agree the phrasing is awkward,
%   but the point we can make is still strong: To make the property easier
%   to enforce, we can slightly weaken its guarantees; the weaker version is
%   a bit more complicated to state but still retains the pragmatic
%%   protections of the original.}\apt{To clarify: Nick's LT\&C policy uses
%   depth-based tags, which do not enforce even our lazy policy. So we need to
%   introduce a version of LT\&C that uses unique activation tags.}

@@

To make our stack safety definitions precise, we make some straightforward
assumptions about
memory layout; our definitions could easily be changed to be parametric over
these assumptions, at the cost of some complexity in their statements.  Memory
consists of an instruction region and a stack that grows upwards. (For simplicity,
we do not consider the heap. Integrating a memory-safety property for the
heap in the style of
\citet{DBLP:conf/post/AmorimHP18} is a natural extension that we discuss
in~\Cref{sec:relwork}.)  The stack has a conventional structure, with
contiguously allocated per-function frames.
The boundary between the frames of a caller and callee is indicated by the $\rsp$ register
at the time of the call.  The stack can be used to
pass arguments or return results in the usual way, by placing them
at the top of the caller's frame. Callers and callees can also communicate
via registers if they wish; indeed, our 
definitions say nothing at all about registers (except $\PCname$ and
$\rsp$). \bcp{This implicitly amounts to choosing a caller-saves register
  convention, though, doesn't it?\apt{I think it is neutral; we could add separate caller register protection if we wanted.}}

Our stack-safety concepts depend critically on the notions of ``call'' and ``return,''
which are not fully explicit in machine code (the same instruction opcode may be
used in some places as part of a call or return sequence and in other places
for other purposes).  We define a call to be
any instruction declared as such in the compiler-provided call structure.
If a call occurs at $\PCname = a$ and $\rsp = s$, then the \emph{corresponding
  return} has occurred when the $\PCname$ points to the next instruction after $a$ and $\rsp = s$. (Our definition
of well-bracketed control flow relates this definition of return to the program's
declared set of return points.)
Our property definitions make no other detailed assumptions about call or return sequences, although
some of the enforcement mechanisms we consider do.

@@

\paragraph*{Software-only policies}
%
Some policies are implemented in software, by modifying code in the
compiler, for example to perform bounds checking~\citep{NagarakatteZMZ09} or
to insert stack canaries~\citep{Cowan+98}. Such approaches can, in principle,
be evaluated in terms of our formal notion of stack safety with a trivial null
policy. But any state maintained by the enforcement mechanism must be in normal
memory and subject to the same confidentiality and integrity concerns (to be
varied, rolled back, etc.), a heavy constraint. Alternately it can be stripped
out as policy state provided that it can be proven to protect itself.

\apt{This still seems wrong to me. The policy part needs somehow to describe everything that isn't varied or rolled back.}
%These approaches are represented as policies that
%accept\bcp{what does it mean for a policy to accept a state??} only initial machine states that are well-formed for the policy (due to
%being produced by the compiler.) The set of policy states is again the unit type
%and the step function constant. Property enforcement comes from the fact that
%only ill-formed states can induce a trace that violates the property.

@@

In this section we examine how an existing enforcement mechanism,
Depth Isolation from \citet{DBLP:conf/sp/RoesslerD18}, can implement
the formal stack-safety property described in the previous section: it in
fact enforces a stronger, inductive version of that property which strengthens the
local state encapsulation component by checking for integrity and
confidentiality violations at every step of the callee's execution trace.

@@

Our micropolicy works as follows. (We assume for ease of exposition that no
arguments are passed on the stack.)
%% \rb{Come back to this later?}.
The micropolicy maintains tags on values in memory of the form $\tagStackDepth{n}$,
indicating locations that belong to the stack frame at activation depth $n$,
or $\tagNoDepth$; it tags the $\PCname$ with tags of the form $\tagPCDepth{n}$.
During normal execution, the micropolicy rules only permit load and store operations
when the target memory is tagged with the same depth as the current {\PCname} tag.
Initially, the entire stack is tagged $\tagNoDepth$,
%\rb{actually, the ``initial frame'' should be tagged with its proper depth}
and the {\PCname} has tag $\tagPCDepth{0}$. These tags are altered
at exactly those points in
the program where the contour of the stack-safety property changes:

@@

There remains the question of how to ensure that the sequences are
invoked at the right places. For a program to enjoy the stack-safety property
(with respect to a particular call map), all we need require is that
the entry sequence be initiated at any instruction marked as a call
in the call map. This is achieved by giving these instructions (another) special tag,
and it is easy to check statically that this has been done correctly.
If the code fails to initiate an exit
sequence at a point where the stack-safety property expects a return,
the micropolicy will incorrectly behave as if execution
is continuing in the callee, but since the callee never has stronger access
rights than the caller, this is harmless.

@@

This conservative policy satisfies not only the stack safety policy of
\cref{sec:lse-and-wbcf},
but also a provably (in Coq) stronger one that enforces local state
encapsulation at every point during execution. To encode that
property, we not only keep track of the execution of one machine
(whose trace we are trying to characterize as safe), but of an entire
stack of machine variations corresponding to nested calls of the
original machine. For each step of the original machine, we take a
step in every variant machine (thus, keeping them in lockstep
execution) and enforce a single-step variant of integrity and
confidentiality tests. In addition, every time a call is made we push
a variation of the original machine state at the time of that call on
top of the variant stack (just like the stack-confidentiality property
enforced trace confidentiality for every call transition), and every time
the machine returns from a call we pop the top variation (to account
for stack confidentiality taking the prefix of the transition trace
until its return).

@@

  Part of the motivation for the strong lockstep stack-safety property
  is that it uncovers errors quickly, and therefore should be quicker
  to test, just like the smarter testing properties
  of~\citet{TestingNI:ICFP}. We have already developed a testing
  framework for debugging our policy enforcement for this property, and
  it would be interesting to explore whether a testing-amenable variant
  of the lazy property also exists.
